---
title: "Is Home Ice Advantage in the NHL Real?"
author: "Everett Robinson"
date: "February 19, 2018"
output: html_document
---

Many people have attempted to answer the question of whether or not home ice advantage exists. A [2013 paper](http://people.stat.sfu.ca/~tim/papers/hca.pdf) by Tim Swartz and Adriano Arce came to the conclusion that home ice advantage in the NHL is real, accounting for about 5% of goals in 2012. Furthermore, they observed that when total goals per game are accoutned for, there is no appreciable change in the home ice advantage over time. Lastly, they performed a one way ANOVA and a pairwise Tukey's HSD test on 16 NHL teams that have played in 30 seasons to determine if there is a significant difference between teams' home ice advantages. The result from their analysis was that there is not sufficient evidence to conclude that home ice advantage varies between teams.

Later in 2017, a [fivethirtyeight blog post](https://fivethirtyeight.com/features/a-home-playoff-game-is-a-big-advantage-unless-you-p lay-hockey/) also tried to tackle this problem, coming to the conclusion that home ice in the regular season amounts to a win percentage of 55.1%, 5.1% better than even odds. In the playoffs this advantage declines slightly to a boost in win percentage that is only 4.8% better on home ice than at a theoretical neutral rink.

The above two results appear fairly consistent with each other, and one might be satisfied that home ice advantage is a real phenomenon. This is probably a safe bet, but I'm all for reproducible research so I'm going to try my hand at answering this question myself and corroborating the previus findings.

There are two questions that I want to attempt to answer:

1. Is there evidence for home ice being advantageous in general? 
2. Is there evidence that this advantage varies from team to team?

I would also like to quantify the extent of home ice advantage in the above analysis, mainly in terms of the expected goal differential it causes. To do all of this I am going to a few different strategies. The first will be a frequentist statistical approach using a paired sample t-test on the home and away teams goals if the assumptions for this test are met. If the assumptions of normality or no outliers are not met, then I will use the  Wilcoxon signed-rank test.

Next, I will investigate the first problem again using a Bayesian approach. To accomplish this, I will use a model based on John Kruschke's [Bayesian estimation supercedes the t-test](http://www.indiana.edu/~kruschke/BEST/BEST.pdf). To do this I will simply use Kruschke's own R package.

To answer the second question, I will implement a one way ANOVA on the collected data to test for sigificant differences in home ice advantage by team.

To start, we will need to get the data for NHL games that have already been played. I will do this using the undocumented NHL statistics API, which returns json files with a mostly self evident structure. These files will be retrieved and parsed using Python and the very easy requests and json libraries. 

```{python, engine = python3}
import requests
import json
import pandas as pd

response = requests.get("https://statsapi.web.nhl.com/api/v1/schedule?startDate=2013-09-18&endDate=2017-08-1")
schedule_json = response.json()

# A function to extract the relevant data from the schedule
# and return it as a pandas dataframe
def extract_game_data(schedule):
    """Given full JSON records for games from the NHL API,
    returns a simplified list of just the data we need.
    """
    games = pd.DataFrame(columns=['date', 
                                  'home_team', 
                                  'home_team_score', 
                                  'away_team', 
                                  'away_team_score'])
    
    for date_obj in schedule_json['dates']:
        date = date_obj['date'];
        for game_obj in date_obj['games']:
            game_type = game_obj['gameType']
            season = game_obj['season']
            home_team_obj = game_obj['teams']['home']
            away_team_obj = game_obj['teams']['away']
            
            home_team = home_team_obj['team']['name']
            home_team_score = home_team_obj['score']
            
            away_team = away_team_obj['team']['name']
            away_team_score = away_team_obj['score']
            
            games = games.append({'date': date,
                                  'season': season,
                                  'game_type': game_type,
                                  'home_team': home_team,
                                  'home_team_score': home_team_score,
                                  'away_team': away_team,
                                  'away_team_score': away_team_score
                                  }, ignore_index=True)
    
    return games
    
games = extract_game_data(schedule_json)
games.to_csv('games.csv', index = False)
```


```{r}
library(tidyverse)
games <- read_csv("games.csv")
```

We will consider theregula r season games that have occured since the last NHL season that experienced a lock-out, the 2012-2013 season, to the end of the last complete season, the 2016-2017 season. This will avoid the current partial solution, which may effect the balance of home vs away games. It will also avoid any disruption and complications that are caused by the introduction of the Vegas Golden Knights, which acquired players from many other teams.

```{r}
reg_season_games <- games %>% 
  filter(game_type == "R")
dim(reg_season_games)
```

Filtering for the above conditions, 4920 regular season games were played across that time span. All teams should have played exactly 164 home games and exactly 164 away games one exception, the Arizona/Phoenix Coyotes. This team underwent a name change to become the Arizona Coyotes between the 2013-2014 and 2014-2015 seasons. To fix this I will simply rename all occurances of the Phoenix Coyotes to the Arizona Coyotes in the data.

```{r}
home_games_by_team <- reg_season_games %>% 
  group_by(home_team) %>% 
  count() %>% 
  rename(home_game_count = n, team = home_team)

away_games_by_team <- reg_season_games %>% 
  group_by(away_team) %>% 
  count() %>% 
  rename(away_game_count = n, team  = away_team)

game_counts_by_team <- home_games_by_team %>% 
  left_join(away_games_by_team) 

game_counts_by_team %>% 
  arrange(home_game_count) %>% 
  head()

game_counts_by_team %>% 
  arrange(home_game_count) %>% 
  tail()
```
```{r}
games$home_team[games$home_team == "Phoenix Coyotes"] = "Arizona Coyotes"
games$away_team[games$away_team == "Phoenix Coyotes"] = "Arizona Coyotes"

reg_season_games <- games %>% 
  filter(game_type == "R")

home_games_by_team <- reg_season_games %>% 
  group_by(home_team) %>% 
  count() %>% 
  rename(home_game_count = n, team = home_team)

away_games_by_team <- reg_season_games %>% 
  group_by(away_team) %>% 
  count() %>% 
  rename(away_game_count = n, team  = away_team)

game_counts_by_team <- home_games_by_team %>% 
  left_join(away_games_by_team) 
game_counts_by_team
```

Some teams are likely inherently stronger than other teams, so we will also want to be sure that the ratio of home to away games between any pair of teams is also close to 1:1. This will ensure that the confounding effect of relative strength is properly controlled for.

```{r, fig.width=8, fig.height=6}
# Get the count of each home_team away_team pair
game_balance_by_pairs <- reg_season_games %>% 
  select(home_team, away_team) %>%
  group_by(home_team, away_team) %>%
  count() %>%
  rename(home_game_count = n)

# Add the total for each matchup, regardless of which team was home or away
game_balance_by_pairs <- game_balance_by_pairs %>%
  mutate(matchup_count = home_game_count + 
                         game_balance_by_pairs[(game_balance_by_pairs$home_team == away_team &
                                                game_balance_by_pairs$away_team == home_team),
                                               'home_game_count'][[1]])

game_balance_by_pairs <- game_balance_by_pairs %>%
  mutate(home_game_proportion = home_game_count / matchup_count)

game_balance_by_pairs %>% ggplot(aes(x = home_team, y = away_team)) + 
  geom_tile(aes(fill = home_game_proportion)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  ggtitle("Proportion of Games Played at Home for All\nPairs of NHL Teams (2012 - 2017)")
```

```{r}
min(game_balance_by_pairs$home_game_proportion)
max(game_balance_by_pairs$home_game_proportion)
```

With a minimum proportion of home games foe each team matchup at 0.4706, and a maximum proportion at 0.5294, it would appear to me that the relative difference in strength between teams is adequately controlled for.
